{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-25 thesis/quicksand.tex
finish quicksand chapter

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{p{0.8\textwidth}}
			\begin{center}
			\begin{tabular}{rlcrl}
				\multicolumn{5}{c}{Initially {\tt int count = \const{0}; mutex\_t m;}} \\
				\\
				& {\bf Thread 1} & & & {\bf Thread 2} \\
				\hline
				11 & \texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & &
						21 & \texttt{atomic\_xadd(\&count, \const{1});} \\
				12 & \texttt{count++;}                                & &
						22 & \texttt{\hilight{olivegreen}{yield}();} \\
				13 & \texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & &
						23 & \texttt{atomic\_xadd(\&count, \const{1});} \\
				14 & \texttt{assert(count >= \const{1});}             & &
						24 & \texttt{assert(count >= \const{2});} \\
			\end{tabular}
			\end{center}
			\\
			(a) Example program with synchronization API calls highlighted.
			\\
			\\
			\begin{center}
			\begin{tabular}{llc}
				{\bf Thread 1} & {\bf Thread 2} & {\bf Value of {\tt count}} \\
				\hline
				\texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & & 0 \\
				\texttt{int tmp = count;} & & 0 \\
							& \texttt{atomic\_xadd(\&count, \const{1});} & 1 \\
							& \texttt{\hilight{olivegreen}{yield}();} & 1 \\
							& \texttt{atomic\_xadd(\&count, \const{1});} & 2 \\
				\texttt{count = tmp + 1;} & & 1 \\
							& \texttt{\hilight{assertfail}{assert(count >= 2);}} & 1 \\
				\texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & & 1 \\
				\texttt{assert(count >= \const{1});} & & 1 \\
			\end{tabular}
			\end{center}
			\\
			(b) Interleaving of (a) in which Thread 2's assertion fails,
			as the single increment of Thread 1 overwrites both those of Thread 2.
		\end{tabular}
	\end{center}
	\caption[Example bug requiring data-race preemption points to expose.]
	{Example bug requiring data-race preemption points to expose.
	Because the two threads use different modes of synchronization to protect their respective accesses to {\tt count},
	preempting on synchronization calls alone is insufficient to expose the bug.
	Rather, Thread 1 must be preempted between its non-atomic load and store of {\tt count}.
	}
	\label{fig:pps-example}
\end{figure}

\begin{figure}[ht!] % TODO: check placement
	\begin{center}
		% FIXME: fix it to be a real tree
		\includegraphics[width=0.9\textwidth]{../oopsla/tree-maximal-only.pdf}
	\end{center}
	\caption[State space of \Cref{fig:pps-example} with synchronization preemption points only.]
	{State space of \Cref{fig:pps-example} with synchronization preemption points only.
	Note that none of these interleavings preempt Thread 1 at the necessary place
	to trigger Thread 2's assertion failure.}
	\label{fig:pps-statespace}
\end{figure}
}

@@

%\subsection{Design}

% TODO: put fig:id here, explain generally multiple state spaces

%Note that
Iterative Deepening is a wrapper algorithm around stateless model checking.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
This chapter
will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

% TODO: fix this section, the way it lists the pps is fucked, figure reference prob dnagling
Iterative Deepening must be seeded with a set of initial state spaces,
which can be any number of subsets of the statically-available preemption points
that prior work model checkers would use.
The upcoming soundness proof relies on the maximal state space being included among these for verification's sake,
but to optimize for finding bugs faster,
implementations may wish to simultaneously to try testing subsets thereof.

For testing user-space code, Quicksand begins with the four
%jobs from \Cref{fig:id}:
\qrevisionminor{possible combinations of preemption points from \Cref{fig:pps-statespace}:}
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,$ $unlock\}$, % FIXME argh
By extension, these also induce preemptions on any other primitives
which use
internal locks,
such as condition variables or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain
Landslide's invariant that only one thread runs between consecutive preemption points,
so the {\tt yield} preemption point is always implicitly enabled.

For kernel-level testing, interrupt-disabling is analogous to locking,
so preemptions must also be introduced
just before a disable-interrupt opcode (on x86, {\tt cli})
and just after interrupts are re-enabled (on x86, {\tt sti}).
During data race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
(note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
\footnote{Some kernels disable preemption without disabling interrupts,
which can be communicated to the model checker using manual annotations,
and must be treated similarly.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}
Quicksand is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
which would result in $2^p$ jobs right off the bat,
although this could potentially be improved in future work (\cref{sec:quicksand-discussion-subsets}).

@@

\begin{figure}[t]
	\begin{center}
        \includegraphics[width=0.9\textwidth]{dr-jobs-v2.pdf}
	\end{center}
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time
		by managing the exploration of multiple state spaces,
		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                %When an access pair is reported as a data race candidate,
                When a data race is reported,
                a new preemption point is added for each involved memory access,
                and new jobs are added for later testing,
                corresponding to different combinations thereof with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

@@

% TODO: fix syntax hilight
% TODO check figure placement
% TODO check figure camption
\begin{figure}[h]
	\begin{center}
	\begin{tabular}{rll}
		& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
		& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
		\\
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		2 & \texttt{\hilight{olivegreen}{free}(x);} \\
		3 & & \texttt{\hilight{commentblue}{// x's memory reallocated}} \\
		4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
		6 & & \texttt{publish(y);} \\
		7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
	\end{tabular}
	\end{center}
	\caption[Reallocation false positive pattern.]
	{Reallocation false positive pattern arising from
	{\tt malloc()} returning the same memory that was just {\tt free()}d.}
	\label{fig:recycle}
\end{figure}

@@

% TODO smae as above
\begin{figure}[h]
	\begin{center}
	\begin{tabular}{rll}
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{publish(x);} & \\
		2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		3 & \texttt{\hilight{olivegreen}{free}(x);} \\
		4 & & \texttt{x2 = get\_published\_x();} \\
		5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
		6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
	\end{tabular}
	\end{center}
	\caption[Reallocation-pattern data race that hides a true bug.]
	{Reallocation-pattern data race that hides a true bug.
	If a single-pass Limited Happens-Before analysis discarded all data race candidates
	with intervening reallocations,
	it would miss the bug in this adversarial program.}
	\label{fig:recycle-bug}
\end{figure}

@@

I present these proofs as they appeared in the OOPSLA paper \cite{quicksand}:
written as sketches in informal prose,
to optimize for rapidly conveying an intuition for why it works
rather than to justify every internal step within the proof structure.%
\footnote{Also because I am already over 200 pages in this thesis, all told.}
% footnote yeah, i'm writing this last
% footnote of footnote yeah, the committee never read this part
A more rigorous treatment
is available in the tech report which accompanied the conference paper \cite{quicksand-soundness}.

@@

\begin{proof}
Any such program must contain an access $a_1$ by one thread T1,
followed by a {\tt free} and a {\tt malloc} possibly by either thread,
followed by an access $a_2$ by the other thread T2,
not depending on the result of the middle malloc.
Without loss of generality, fix T1 to perform the {\tt free} and T2 the subsequent {\tt malloc},
as shown in \Cref{fig:recycle-bug}.
The other cases are similar,
although note that if T2 performs the {\tt free},
and the {\tt malloc} is reordered before it,
T2's final access will be a use-after-free immediately.
Let us also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
hence, there must also be some ``publish'' action $p$ by T1 which communicates the address to T2.
Because this is a true potential data race,
$p$ must occur before $a_1$, as $a_2$ cannot be reordered before $p$.

@@

With this preemption point,
DPOR will reorder $a_2$ before $a_1$, while not changing $a_2$'s location.
As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
Hence $a_1$ and $a_2$ %will be in the same allocation;
can race without appearing to fall under the reallocation pattern.
\end{proof}

This spells QED so we are done \cite{vargomax}. % even more :)))
Note that this proof does not rely on the existence of preemption points on
the internal lock of {\tt malloc} or {\tt free},
which is an ideal candidate to ignore via {\tt without\_function} (\cref{sec:landslide-pps})
to reduce state space size.
Future work may generalize this proof structure to not rely on specific knowledge
of {\tt malloc()}'s and {\tt free()}'s behaviour,
but instead to require only any intervening synchronization event,
thereby extending the overall soundness proof to accomodate Pure Happens-Before as well as Limited Happens-Before.
The experiments in future chapters of this thesis will assume that this holds.

@@

The available command-line options for configuring its
CPU-time budget, exploration modes, and so on
are listed in \cref{sec:landslide-quicksand-options}.
Additionally, Quicksand periodically issues a progress report
at fixed intervals to inform the user on the completion, bug-finding, and/or estimated progress of each job.
\Cref{fig:quicksand-progress} shows an example.
I highlight a few notable features of the jobs therein
to serve as a concrete example that may cement the reader's intuition
of \cref{sec:quicksand-id}'s more abstract algorithm descriptions:
\begin{itemize}
	\item Jobs 0, 1, 2, and 3 are the initially-seeded state spaces (\cref{sec:quicksand-initial-pps}).
	\item Job 2 reports a bug found, and shows the filename of the HTML preemption trace
		(\cref{sec:landslide-bugreport}, \Cref{fig:bugreport})
		which the user should examine to diagnose it.%
		\footnote{I actually cheated by copy/pasting this job alone from a different run of Quicksand;
		the other jobs come from a test with the bug already fixed,
		in order that exploration progress far enough to defer some jobs for the sake of example.
		In a real execution, the superset jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
	\item Jobs 4 and 6 are the ``small'' jobs added to test the two data races in isolation;
		5 and 7 are the corresponding ``large'' jobs (\cref{sec:quicksand-dr-pps}).
	\item Job 11 is the maximal state space, containing all synchronization preemption points
		and both currently-known data races.
	\item The intermediate jobs 8, 9, and 10 have been suspended for having ETAs
		at least twice as large as the provided CPU budget (1 hour),
		according to the ETA scaling factor heuristic (\cref{sec:quicksand-heuristics}).
	\item Note that job 11's ETA is currently lower than 8's, 9's, and 10's, despite being a strict superset of each.
		This corresponds to the surprising ETA inversion situation discussed in \cref{sec:quicksand-choosing}:
		Quicksand simply hasn't made as much progress into job 11 (compare their percentage estimates rather than ETAs)
		for its ETA to be accurate enough yet.
\end{itemize}
Future work could extend these progress reports to be more interactive,
allowing the user to reprioritize state spaces at her whim
or to disable certain data-race preemption points after checking them by hand,
as discussed in \cref{sec:future-friendly}.

% TODO: syn hi
\begin{figure}[t]
	\begin{center}
		\footnotesize
	\begin{tabular}{l}
			% version from buggy run
		%\texttt{==== PROGRESS REPORT ====} \\
		%\texttt{total time elapsed: 40s} \\
		%\texttt{[JOB 0] COMPLETE (4 interleavings tested; 7s elapsed)} \\
		%\texttt{       PPs: \{ \}} \\
		%\texttt{[JOB 1] Running (10.677083\%; ETA 2m 15s)} \\
		%\texttt{       Log: id/ls-output.log.ujrk9U -- PPs: \{ 'mutex\_lock' \}} \\
		%\texttt{[JOB 2] BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested; 1 preemptions)} \\
		%\texttt{       Log of lprintf()s: id/ls-setup.log.7UMGLr} \\
		%\texttt{       Log: id/ls-output.log.3Ol8rG -- PPs: \{ 'mutex\_unlock' \}} \\
		%\texttt{[JOB 3] CANCELLED} \\
		%\texttt{       PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}} \\
		%\texttt{[JOB 5] COMPLETE (4 interleavings tested; 7s elapsed)} \\
		%\texttt{       PPs: \{ 'data race 5@ 0x1000ec5' \}} \\
		%\texttt{[JOB 13] Setting up...} \\
		%\texttt{       PPs: \{ 'data race 6@ 0x1000ec5' \}} \\
		%\texttt{[JOB 14] Pending...} \\
		%\texttt{       PPs: \{ 'mutex\_unlock' 'data race 6@ 0x1000ec5' \}} \\
		%\texttt{=========================} \\

		\texttt{==== PROGRESS REPORT ====} \\
		\texttt{total time elapsed: 2m 52s} \\
		\texttt{[JOB 0] COMPLETE (4 interleavings tested; 7s elapsed)} \\
		\texttt{~~~~PPs: \{ \}} \\
		\texttt{[JOB 1] Running (21.932870\%; ETA 8m 14s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_lock' \}} \\
		% version that came from the actual test
		%\texttt{[JOB 2] Running (46.373457\%; ETA 4m 35s)} \\
		%\texttt{~~~~ PPs: \{ 'mutex\_unlock' \}} \\
		\texttt{[JOB 2] BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested)} \\
		%\texttt{~~~~Log of lprintf()s: id/ls-setup.log.7UMGLr} \\
		\texttt{~~~~ PPs: \{ 'mutex\_unlock' \}} \\
		\texttt{[JOB 3] Running (9.852431\%; ETA 24m 32s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}} \\
		\texttt{[JOB 4] COMPLETE (6 interleavings tested; 9s elapsed)} \\
		\texttt{~~~~PPs: \{ 'data race @ 0x102917' \}} \\
		\texttt{[JOB 5] Running (3.710938\%; ETA 1h 2m 14s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' \}} \\
		\texttt{[JOB 6] COMPLETE (4 interleavings tested; 8s elapsed)} \\
		\texttt{~~~~PPs: \{ 'data race @ 0x1000ecf' \}} \\
		\texttt{[JOB 7] Running (6.119792\%; ETA 33m 14s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x1000ecf' \}} \\
		\texttt{[JOB 11] Running (3.670247\%; ETA 50m 16s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' 'data race @ 0x1000ecf' \}} \\
		\texttt{[JOB 8] Deferred... (33.340567\%; ETA 2h 6m 3s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_unlock' 'data race @ 0x102917' \}} \\
		\texttt{[JOB 9] Deferred... (34.466226\%; ETA 2h 35m 37s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_unlock' 'data race @ 0x1000ecf' \}} \\
		\texttt{[JOB 10] Deferred... (11.113790\%; ETA 4h 20m 31s)} \\
		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'data race @ 0x102917' \}} \\
		% no, i deleted all of them pretending the other data races dont exist
		%\texttt{And 63 more pending jobs should time allow.} \\
		\texttt{=========================} \\
	\end{tabular}
	\end{center}
	\caption{Example Quicksand progress report showing the various possible job states.}
	\label{fig:quicksand-progress}
\end{figure}

>>>>

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{p{0.9\textwidth}}
			\begin{center}
			\begin{tabular}{rlcrl}
				\multicolumn{5}{c}{Initially {\tt int count = \const{0}; mutex\_t m;}} \\
				\\
				& {\bf Thread 1} & & & {\bf Thread 2} \\
				\hline
				11 & \texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & &
						21 & \texttt{atomic\_xadd(\&count, \const{1});} \\
				12 & \texttt{count++;}                                & &
						22 & \texttt{\hilight{olivegreen}{yield}();} \\
				13 & \texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & &
						23 & \texttt{atomic\_xadd(\&count, \const{1});} \\
				14 & \texttt{assert(count >= \const{1});}             & &
						24 & \texttt{assert(count >= \const{2});} \\
			\end{tabular}
			\end{center}
			\\
			\begin{center}
			(a) Example program with synchronization API calls highlighted.
			\end{center}
			\\
			\begin{center}
			\begin{tabular}{llllc}
				& {\bf Thread 1} & & {\bf Thread 2} & {\bf Value of {\tt count}} \\
				\hline
				11 & \texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & & & 0 \\
				12a & \texttt{int tmp = count;} & & & 0 \\
							& & 21 & \texttt{atomic\_xadd(\&count, \const{1});} & 1 \\
							& & 22 & \texttt{\hilight{olivegreen}{yield}();} & 1 \\
							& & 23 & \texttt{atomic\_xadd(\&count, \const{1});} & 2 \\
				12b & \texttt{count = tmp + 1;} & & & 1 \\
				13 & \texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & & & 1 \\
				14 & \texttt{assert(count >= \const{1});} & & & 1 \\
							& & 24 & \texttt{\hilight{assertfail}{assert(count >= 2);}} & 1 \\
			\end{tabular}
			\end{center}
			\\
			%(b) Interleaving of (a) in which Thread 2's assertion fails,
			%as the single increment of Thread 1 overwrites both those of Thread 2.
			(b) Buggy interleaving of (a) in which
			the single increment of Thread 1 overwrites both those of Thread 2.
			Note the data-race preemption at 12a-12b.
		\end{tabular}
	\end{center}
	\caption[Example bug requiring data-race preemption points to expose.]
	{Example bug requiring data-race preemption points to expose.
	Because the two threads use different modes of synchronization to protect their respective accesses to {\tt count},
	preempting on synchronization calls alone is insufficient to expose the bug.
	Rather, Thread 1 must be preempted between its non-atomic load and store of {\tt count}.
	}
	\label{fig:pps-example}
\end{figure}

\begin{figure}[ht!] % TODO: check placement
	\begin{center}
		\includegraphics[width=0.98\textwidth]{dr-statespace.pdf}
	\end{center}
	\caption[State space of \Cref{fig:pps-example} with synchronization preemption points only.]
	{State space of \Cref{fig:pps-example} with synchronization preemption points only.
	Note that none of these interleavings preempt Thread 1 at the necessary place (between 12a and 12b)
	to trigger Thread 2's assertion failure.}
	\label{fig:pps-statespace}
\end{figure}
}

@@

%Note that
Iterative Deepening is a wrapper algorithm around stateless model checking.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
(\cref{sec:landslide-dpor})
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
This chapter
will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

Iterative Deepening must be seeded with a set of initial state spaces,
which can be any number of subsets of the statically-available preemption points
that prior work model checkers would use.
The upcoming soundness proof relies on the maximal state space being included among these for verification's sake,
but to optimize for finding bugs faster,
implementations may wish to simultaneously to try testing subsets thereof.

For testing user-space code, Quicksand begins with the four
%jobs from \Cref{fig:id}:
\qrevisionminor{possible combinations of preemption points from \Cref{fig:pps-statespace}:}
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,$ $unlock\}$, % FIXME argh
By extension, these also induce preemptions on any other primitives
which use
internal locks,
such as condition variables or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain
Landslide's invariant that only one thread runs between consecutive preemption points,
so the {\tt yield} preemption point is always implicitly enabled.
%
For kernel-level testing, interrupt-disabling is analogous to locking,
so preemptions must also be introduced
just before a disable-interrupt opcode (on x86, {\tt cli})
and just after interrupts are re-enabled (on x86, {\tt sti}).
During data race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
(note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
\footnote{Some kernels disable preemption without disabling interrupts,
which can be communicated to the model checker using manual annotations,
and must be treated similarly.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}
Quicksand is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
which would result in $2^p$ jobs, %right off the bat,
although this could potentially be improved in future work (\cref{sec:quicksand-discussion-subsets}).

@@

\begin{figure}[t]
	\begin{center}
        \includegraphics[width=0.9\textwidth]{dr-jobs-v3.pdf}
	\end{center}
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time
		by managing the exploration of multiple state spaces,
		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                %When an access pair is reported as a data race candidate,
                When a data race is reported,
                a new preemption point is added for each involved memory access,
                and new jobs are added for later testing,
                corresponding to different combinations thereof with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

@@

\begin{figure}[h]
	\begin{center}
	\begin{tabular}{rll}
		& \multicolumn{2}{c}{\texttt{\ctype{struct x} \{ \ctype{int} foo; \ctype{int} baz; \} *x;}} \\
		& \multicolumn{2}{c}{\texttt{\ctype{struct y} \{ \ctype{int} bar; \} *y;~~~~~~~~~}} \\
		\\
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{\hilight{assertfail}{x->foo = ...;}} & \\
		2 & \texttt{\call{free}(x);} \\
		3 & & \texttt{\ccomment{// x's memory reallocated}} \\
		4 & & \texttt{y~=~\call{malloc}(\flow{sizeof} *y);} \\
		5 & & \texttt{\ccomment{// ...initialize...}}\\
		6 & & \texttt{\call{publish}(y);} \\
		7 & & \texttt{\hilight{assertfail}{y->bar = ...;}} \\
	\end{tabular}
	\end{center}
	\caption[Reallocation false positive pattern.]
	{Reallocation false positive pattern arising from
	{\tt malloc()} returning the same memory that was just {\tt free()}d.}
	\label{fig:recycle}
\end{figure}

@@

\begin{figure}[h]
	\begin{center}
	\begin{tabular}{rll}
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{\call{publish}(x);} & \\
		2 & \texttt{\hilight{assertfail}{x->foo = ...;}} & \\
		3 & \texttt{\call{free}(x);} \\
		4 & & \texttt{x2 = \call{get\_published\_x}();} \\
		5 & & \texttt{\ccomment{// x's memory recycled}} \\
		6 & & \texttt{y~=~\call{malloc}(\flow{sizeof} *y);} \\
		7 & & \texttt{\hilight{assertfail}{x2->foo = ...;}} \\
	\end{tabular}
	\end{center}
	\caption[Reallocation-pattern data race that hides a true bug.]
	{Reallocation-pattern data race that hides a true bug.
	If a single-pass Limited Happens-Before analysis discarded all data race candidates
	with intervening reallocations,
	it would miss the bug in this adversarial program.}
	\label{fig:recycle-bug}
\end{figure}

@@

I present these proofs as they appeared in the OOPSLA paper \cite{quicksand}:
written as sketches in informal prose,
to optimize for rapidly conveying an intuition for why it works
rather than to justify every internal step within the proof structure.%
\footnote{Also because this thesis is long enough already.}
% footnote yeah, i'm writing this last
% footnote of footnote yeah, the committee never read this part
A more rigorous treatment
is available in the tech report which accompanied the conference paper \cite{quicksand-soundness}.

@@

\begin{proof}
Any such program must contain an access $a_1$ by one thread T1,
followed by a {\tt free()} and a {\tt malloc()} possibly by either thread,
followed by an access $a_2$ by the other thread T2,
not depending on the result of the middle malloc.
Without loss of generality, fix T1 to perform the {\tt free()} and T2 the subsequent {\tt malloc()},
as shown in \Cref{fig:recycle-bug}.
The other cases are similar,
although note that if T2 performs the {\tt free()},
and the {\tt malloc()} is reordered before it,
T2's final access will be a use-after-free immediately.
Let us also assume the only way for the program to get pointers to heap memory is through {\tt malloc()};
hence, there must also be some ``publish'' action $p$ by T1 which communicates the address to T2.
Because this is a true potential data race,
$p$ must occur before $a_1$, as $a_2$ cannot be reordered before $p$.

@@

With this preemption point,
DPOR will reorder $a_2$ before $a_1$, while not changing $a_2$'s location.
As T2's {\tt malloc()} now occurs before T1's {\tt free()}, it will allocate different memory.
Hence $a_1$ and $a_2$ %will be in the same allocation;
can race without appearing to fall under the reallocation pattern.
\end{proof}

This spells QED so we are done \cite{vargomax}. % even more :)))
Note that this proof does not rely on the existence of preemption points on
the internal lock of {\tt malloc()} or {\tt free()},
which is an ideal candidate to ignore via {\tt without\_function} (\cref{sec:landslide-pps})
to reduce state space size.
Future work may generalize this proof structure to not rely on specific knowledge
of {\tt malloc()}'s and {\tt free()}'s behaviour,
but instead to require only any intervening synchronization event,
thereby extending the overall soundness proof to accomodate Pure Happens-Before as well as Limited Happens-Before.
The experiments in future chapters of this thesis will assume that this holds.

@@

The available command-line options for configuring its
CPU-time budget, exploration modes, and so on
are listed in \cref{sec:landslide-quicksand-options}.
Additionally, Quicksand periodically issues a progress report
at fixed intervals to inform the user on the completion, bug-finding, and/or estimated progress of each job.
\Cref{fig:quicksand-progress} shows an example.
I highlight a few notable features of the jobs therein
to serve as a concrete example that may cement the reader's intuition
of \cref{sec:quicksand-id}'s more abstract algorithm descriptions:

\begin{figure}[t]
	\begin{center}
		\footnotesize
	\begin{tabular}{l}
			% version from buggy run
		%\texttt{==== PROGRESS REPORT ====} \\
		%\texttt{total time elapsed: 40s} \\
		%\texttt{[JOB 0] COMPLETE (4 interleavings tested; 7s elapsed)} \\
		%\texttt{       PPs: \{ \}} \\
		%\texttt{[JOB 1] Running (10.677083\%; ETA 2m 15s)} \\
		%\texttt{       Log: id/ls-output.log.ujrk9U -- PPs: \{ 'mutex\_lock' \}} \\
		%\texttt{[JOB 2] BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested; 1 preemptions)} \\
		%\texttt{       Log of lprintf()s: id/ls-setup.log.7UMGLr} \\
		%\texttt{       Log: id/ls-output.log.3Ol8rG -- PPs: \{ 'mutex\_unlock' \}} \\
		%\texttt{[JOB 3] CANCELLED} \\
		%\texttt{       PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}} \\
		%\texttt{[JOB 5] COMPLETE (4 interleavings tested; 7s elapsed)} \\
		%\texttt{       PPs: \{ 'data race 5@ 0x1000ec5' \}} \\
		%\texttt{[JOB 13] Setting up...} \\
		%\texttt{       PPs: \{ 'data race 6@ 0x1000ec5' \}} \\
		%\texttt{[JOB 14] Pending...} \\
		%\texttt{       PPs: \{ 'mutex\_unlock' 'data race 6@ 0x1000ec5' \}} \\
		%\texttt{=========================} \\

		\texttt{\ccomment{==================================== PROGRESS REPORT ====================================}} \\
		\texttt{\flow{total time elapsed: 2m 52s}} \\
		\texttt{\ccomment{[JOB 0]} \ctype{COMPLETE (4 interleavings tested; 7s elapsed)}} \\
		\texttt{~~~~\ccomment{PPs: \{ \}}} \\
		\texttt{\ccomment{[JOB 1]} \flow{Running (21.932870\%; ETA 8m 14s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' \}}} \\
		% version that came from the actual test
		%\texttt{\ccomment{[JOB 2]} Running (46.373457\%; ETA 4m 35s)} \\
		%\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' \}}} \\
		\texttt{\ccomment{[JOB 2]} \hilight{assertfail}{BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested)}} \\
		%\texttt{~~~~Log of lprintf()s: id/ls-setup.log.7UMGLr} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' \}}} \\
		\texttt{\ccomment{[JOB 3]} \flow{Running (9.852431\%; ETA 24m 32s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}}} \\
		\texttt{\ccomment{[JOB 4]} \ctype{COMPLETE (6 interleavings tested; 9s elapsed)}} \\
		\texttt{~~~~\ccomment{PPs: \{ 'data race @ 0x102917' \}}} \\
		\texttt{\ccomment{[JOB 5]} \flow{Running (3.710938\%; ETA 1h 2m 14s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' \}}} \\
		\texttt{\ccomment{[JOB 6]} \ctype{COMPLETE (4 interleavings tested; 8s elapsed)}} \\
		\texttt{~~~~\ccomment{PPs: \{ 'data race @ 0x1000ecf' \}}} \\
		\texttt{\ccomment{[JOB 7]} \flow{Running (6.119792\%; ETA 33m 14s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x1000ecf' \}}} \\
		\texttt{\ccomment{[JOB 11]} \flow{Running (3.670247\%; ETA 50m 16s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' 'data race @ 0x1000ecf' \}}} \\
		\texttt{\ccomment{[JOB 8]} \call{Deferred... (33.340567\%; ETA 2h 6m 3s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' 'data race @ 0x102917' \}}} \\
		\texttt{\ccomment{[JOB 9]} \call{Deferred... (34.466226\%; ETA 2h 35m 37s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' 'data race @ 0x1000ecf' \}}} \\
		\texttt{\ccomment{[JOB 10]} \call{Deferred... (11.113790\%; ETA 4h 20m 31s)}} \\
		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'data race @ 0x102917' \}}} \\
		% no, i deleted all of them pretending the other data races dont exist
		%\texttt{And 63 more pending jobs should time allow.} \\
		\texttt{\ccomment{=========================================================================================}} \\
	\end{tabular}
	\end{center}
	\caption{Example Quicksand progress report showing the various possible job states.}
	\label{fig:quicksand-progress}
\end{figure}
